---
title: "Introduction to ldc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ldc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ldc)
```

The goal of ldc is to provide functions for quickly creating pollutant
load duration curves (LDCs). LDCs are an extension of flow duration curves 
(FDCs). The FDC is essentially a cumulative distribution plot with observed
streamflow values on the y-axis, and the proportion of values that exceed 
a given flow value on the x-axis. Specifically, the FDC displays the exceedance 
probability $p$ on the x-axis and the associated discharge ($Q$) on the y-axis
[@vogel_flowduration_1994]. Typically, mean daily discharge is used to
calculate the flow exceedance values. Using this approach, flows can be
categorized into flow duration intervals such as "Flood Flows," "High Flows,"
"Moderate Flows," and "Low Flows." 

The LDC is developed by multiplying allowable pollutant concentrations by the
daily streamflow volume to identify allowable pollutant loads across flow
duration intervals. Measured pollutant concentrations can be added to the
duration curve by multiplying the concentrations and streamflow volume to derive
an instantaneous load at a given exceedance percentile. By overlaying measured
values over the LDC line, we can develop some inference for what conditions
pollutant concentrations and loads exceed allowable concentrations. 
@cleland_tmdl_2003 and @morrison_development_2008 provide additional information
for LDC development and use in pollutant loading assessments.

# Input Data

ldc requires a dataframe with at least two columns, discharge ($Q$) and
measured pollutant concentrations ($C$). It is assumed, $Q$ is mean daily
discharge, any units can be used (cfs, cms, etc.) $C$ can be any measured 
pollutant concentration associated with the value of $Q$ on the same row. The
package includes a dataset, `tres_palacios`, that demonstrates how this is 
formatted.

```{r importdata}
## load required packages
library(dplyr)

## load example data
df <- as_tibble(tres_palacios) %>%
  ## filter this to the last 5 years of data
  filter(Date >= as.Date("2015-01-01"))

## show the data filtered to paired Q and C observations
df %>% filter(!is.na(Indicator_Bacteria))


```

ldc also utilizes the [units](https://cran.r-project.org/package=units) package
to handle unit conversions. Therefore, $Q$ and $C$ will need to be formatted as
units objects before using any ldc functions. In the example below, we need to
create a new unit, "cfu", for the Indicator_Bacteria variable and change both 
Flow and Indicator_Bacteria to unit objects.

```{r formatdata}
## load required packages
library(units)


## make the cfu unit
install_unit("cfu")

## change Q and C to unit objects with appropriate units
df <- df %>%
  mutate(Flow = set_units(Flow, "ft^3/s"),
         Indicator_Bacteria = set_units(Indicator_Bacteria, "cfu/100mL"))
df
```

Now the dataframe is formatted for use in ldc functions. You can see the unit
objects include the appropriate units when printed. 

# Functions

ldc is composed of three major functions:

  - `calc_ldc` takes an input dataset of matched flow and pollutant 
  concentrations to generate a dataframe with exceedance probabilities for
  allowable pollutant loads (the LDC), measured pollutant concentrations
  converted to loads, alongside user specified flow duration intervals.
 
 - `summ_ldc` uses the output from `calc_ldc` to generate a summary dataframe
 grouped by flow duration intervals.
 
 - `draw_ldc` uses the output from both functions to generate an LDC figure as 
 a ggplot object.
 

**`calc_ldc`**

The `calc_ldc` function is the first function to run. It has six arguments:

 - .tbl is the dataframe formatted as described above.
 
 - Q is the variable name in .tbl for discharge or flow.
 
 - C is the variable name for measured pollutant concentration.
 
 - allowable_concentration is an object of class units specifying the allowable
 pollutant concentration (water quality standard). This must have the same units
 as C.
 
 - breaks is a numeric vector from (1 -> 0) that indicates the break points for
 flow duration intervals. The length of the vector should be 1 more than the number of
 flow duration intervals. For example, `c(1, 0.9, 0.6, 0.4, 0.1, 0)` has 6 breaks and
 5 flow duration intervals.
 
 - labels is a vector with the names for each flow duration interval. For
 example, `c("Highest flows", "High Flows", "Moderate Flows", "Dry Conditions", "Lowest Flows")`.
 

```{r calcldc}
## set the allowable concentration
allowable_concentration <- 126
units(allowable_concentration) <- "cfu/100mL"

## calculate the ldc
df_ldc <- calc_ldc(df,
                   Q = Flow,
                   C = Indicator_Bacteria,
                   allowable_concentration = allowable_concentration,
                   breaks = c(1, 0.9, 0.6, 0.4, 0.1, 0),
                   labels = c("Highest flows", "High Flows", "Moderate Flows", "Dry Conditions", "Lowest Flows"))

## show the result
df_ldc %>% filter(!is.na(Indicator_Bacteria))
```

**`summ_ldc`**

The `summ_ldc` function is the second function to run. It has six arguments:

 - .tbl is a dataframe, preferably the output from `calc_ldc()` or formatted 
 exactly like it.

 - Q is the variable name in .tbl for discharge or flow.

 - C is the variable name in .tbl for measured pollutant concentration.

 - Exceedance is the variable name in .tbl for flow exceedance probabilities. 
 This defaults to the output created by `calc_ldc()` so in most cases you can 
 leave it alone.
 
 - groups is the variable name in .tbl with the flow interval names. This
 defaults to the output created by `calc_ldc` so in most cases you can leave it 
 alone.
 
 - method is a string describing the method desired for summarizing pollutant
 concentration. It can be one of `"geomean"`, `"mean"`, or `"median"`. 
 
```{r calcsumm}
df_sum <- summ_ldc(df_ldc, 
                   Q = Flow, 
                   C = Indicator_Bacteria, 
                   Exceedance = P_Exceedance,
                   groups = Flow_Category,
                   method = "geomean")
df_sum
```

**`draw_ldc`**

The final function is `draw_ldc` which uses the output from the previous two
functions to generate a plot as a ggplot object. There are many arguments in 
this function, I won't describe them all but details are available using 
`?draw_ldc`. Importantly, since the output is a ggplot object, the output can
be modified using many of the available ggplot functions.

```{r draw}
library(ggplot2)

p1 <- draw_ldc(df_ldc, 
               df_sum, 
               y_lab = expression(paste(italic("E. coli"))),
               ldc_legend_name = "LDC at 35cfu/100mL",
               measurement_name = "Measured Values",
               summary_name = "Summarized Loads (Geometric Mean)",
               label_nudge_y = log10(1000)) + 
  scale_y_log10() +
  annotation_logticks(sides = "l") +
  theme_bw() +
  theme(legend.title = element_blank(),
        legend.direction = "vertical",
        legend.position = "bottom")

p1
```

